Writing Trash Bin

### 2024-03-22 paper.qmd
finished rewrite

Unlike uncertainty, the concept of the "information" in a plot is not entirely avoided and works such as *The Grammar of Graphics* attempt to outline exactly what information is contained within a plot. When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not, this pipeline is illustrated in @fig-pipeline, and contains all the steps of the grammar of graphics [@Leland2005]. First the mathematical information in a plot is established in the "varset" setps, and then that mathematical information is visually represented, which is defined in the "graph" steps. Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical. However because uncertainty is not defined mathematically, we cannot reliably add it in at the "statistic" stage. Additionally, the grammar of graphics was not designed to extend to "uncertainty visualisations" as @Leland2005 referred to uncertainty as "semantics", so concepts that are common to uncertainty mathematically, such as resampling do not quite fit within the framework. For example, if we have one set of data and two graphics, where one graphic represents the data as a series of points, and the the other depicts a smoothed density using a line. These two plots differ in *both* the statistic and geometry stages of the grammar of graphics, however as n (the number of observations in our sample) gets larger, these two plots will become visually indistinguishable (depending on the collision modifier used). This result aligns nicely with large sample theory, because as n increases the data and mass should become mathematically indistinguishable, and many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness so this result is likely also practically true [@Kale2018; @Newburger2022; @Hofmann2012] but despite the increasing similarity in what is actually depicted in the graphic, the sample and mass visualisations would remain different visualisations within the grammar of graphics framework. Additionally, authors rarely attribute this visual version of large sample theory to results where it is clearly applicable. @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result. The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here. Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these graphics differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear.

@@

# Uncertainty is noise, not signal
Uncertainty is noise, but most evaluation experiments measure it as signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it, if that message is invalid.  When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

@@

### Heuristic checks
- Have a better hypothesis? Like its more specific. They are looking for a very specific thing *wrong* with a visualisation.
- We wary of trying to control something and observe it at the same time

A lot of work that identifies heuristics or biases in plots do not hinge on the previous assumptions. This work also provides useful insights for experiments by highlights pitfalls participants might fall into when they review the results of evaluation experiments [@Hullman2016]. @Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars. Similar effects have been identified in other types of uncertainty displayed. @Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution. The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015]

### Just noticiable signal
Alternatively visualisation research could shift away from the accuracy concept all together and either allow for open ended responses, or ask questions that are more abstract to try and better replicate the large structural understanding that comes with the insights gained from visualisation.

### Thinking "smaller picture" with
Many visualisation experiments try to compare two plots with *several* differences, but do not seem to be interested in the *mechanisms* by which we extract information from visualisations. Let us consider a decision making experiment, there are several steps that Simply asking "Do people make better decisions with

**Experiment rant from confirmation**
The field could take a more "exploratory" approach to experimentation and ask multiple accuracy questions for multiple visualisations with different visual encodings and try to identify the *trade-off* in the information a particular plot conveys. Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations. This experimental framework could replicate this process.

### Comparisons to lineup plots and hypothesis testing
It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot]. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.

![Difference between the null plot vs the uncertainty plot.](hypvs.jpeg){#fig-hypvs}

The problem of trying to assess uncertainty without knowing *why* we are depiting uncertainty has been noticed by other authors. @Spiegelhalter2017 noted we "cannot assess the quality of risk communication unless the objectives are clear".

# Great Examples 
(good decision making example) @Boone2018 organised responses from participants by the approach they used and tried to account for participants prior beliefs (the method of eliciting prior beleifs was, however, questionable).

> Edit notes: this section seems to go off a bit on dimensionality of visual tasks and stuff that might be better put in the noise and signal section about measuring uncertainty.

Of course we do not believe every uncertainty visualisation is unusable and every experiment is build upon misguided assumptions. Despite the common problems detialled in the previous sections, there is some interesting work in the uncertainty visualisation space. This can come in the form of an uncertainty visualisation that identifies or fixes pitfalls in the previous sections (without identifying them) or does an experiment that identifies useful information that avoids these pitfalls.

Uncertainty that is created by steps that are further down the pipeline are often ignored by the uncertainty visualisation literature, however there are visualisations that make these considerations. If uncertainty had a more encapsulating mathematical definition, the uncertainty created at earlier stages might be able to be visualised with the model uncertainty that is typically expressed through confidence intervals. That being said, even with a definition that allows us to combine these uncertainties into a single value, it still might be desirable to see which stage of the analysis is creating uncertainty in the visualisaiton. One example of a visualisation that includes that information is visualisations of climate scenario uncertainty, as shown in @fig-climatescenario. Uncertainty is introduced into this model by the choices of individuals which is difficult to model or predict and is also unlikely to be constant over time. Therefore specific scenarios are modelled and shown, each with their own model uncertainty. This highlights a method that can be used to visualise uncertainty from earlier in the pipeline that is often forgotten or ignored via assumptions.

There are also some interesting qualitative studies that take an investigative stance to the use of uncertainty in decision making rather than an evaluation stance. @Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making. This study was qualitative and rather than setting a "correct" answer that would have forced a value of risk on the participants, the experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty. This study was an excellent example in a useful experimental design. They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations.

![I will just make my own version of a climate scenario plot, this one is just here from a random site as a place holder](climatescenario.png){fig-climatescenario}

Another method that has been used is to visualise samples from the distribution we are trying to draw inference on instead of estimates along with a variance or error. This has been used in maps with the pixel map [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000] . The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap and the New York Times class mobility figure, shown in @fig-nyt, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).  Therefore, so long as the sample size of the number of outcomes is kept to a level where individual samples are still visible, there is no evidence that an outcome plot *needs* to be animated. Visualisations that opt to express a signal as a sample rather than an estimate have the potential to supress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments. This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing. These sampling methods can show the messiness of the data that sits behind a model.

@@

A final method to perform signal supression is simply to visualise the data if it is available. An example of this is shown in @fig-census for racial distributions spatially in america, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/). This map shows the typical causes of uncertainty in a spatial model, i.e. regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc, but it avoids the need to create a visualisation with a specific signal in mind. This is the technique typically employed by exploratory data analysis, which means it's lack of a specified signal means there is both *no* uncertainty (since we are technically not performing inference) but in the event we *do* implicitly perform inference, there is some hedging. In this sense, the best uncertainty visualisation you can get without specifying a signal you want to convey is visualising the data itself. The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out. The raw data can also be used in a similar method to the HOPs plot, where a statistic of importance is implied by the visualisation, but the signal is supressed using the data itself instead of resampling methods. This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through. @Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the lineup protocol was implemented.

@@

These great examples show there are visualisations that are being made that have a good conceptualisation of uncertainty and express it as signal supression, however this work is not being done formally within visualisation. These industry visualisation are likely made with a large amount of back and forth between different levels of an organisation, something that is important for good visual communication [@Spiegelhalter2017].

@@

Our first suggestion for future work is a mathematical defintion of uncertainty. Specifically, a definition that has a place for all the concepts that currently sit under the "uncertainty" umbrella, explains how these concepts relate to uncertianty, and also allows us to quantify them in a way where they can be combined or visualised. This work would not only benefit uncertainty visualisaiton, but also statistics more broadly. There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately. Elements such as imputed data, assumptions, sampling methods, and analyst decisions all have their own independent work, quantification and discussions, but methods to brand between these concepts are few and far between. A conceptual framework that allows us to combine these results would be incredibly beneficial to the field.

The concept of uncertainty should also be formalised within the grammar of graphics. Not only would a formalisation of uncertainty visualisation within the grammar of graphics allow us to iron out some of these confusions, but they would make it easier to understand when existing visualisation methods apply and can be used to explain our results.

```{r}
#| include: false

# put a visualisation of a dot plot with increasing sample size here to show my point
```

If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal supression in an experimental design. It is clear that value retrival and questions with a specific ground truth are not appropriate, as uncertainty visualisation will simply follow the existing visualisation rules. Testing a secondary feature of a plot, that is, how strong a signal is becomes a little more complicated. Line up protocols have been used in adjacent work that look at the strength of the signal in a plot, and this idea of identifying if plots have some "barely noticiable differences" could be utilised. There is also the possibility that uncertainty visualisation evaluations will need to swap to a qualitative methodology where participants are allowed to freely comment on what they notice in graphics until we establish how the existence of noise can be observed.

Additionally, information visualisation needs a more precise concept of the difference between two plots, and that framework should be utilised in experiments. Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results. Visualisation has the grammar of graphics, however we have already discussed how this can break down in the case of distribution visualisation.

If a visualisation researcher would prefer to perform experiments rather than formalise methods, the task dependency many authors in uncertainty visualisation mention would be a useful direction for research. Unfortunately the current visualisation evaluation literature is *far* too noisy to allow us to identify what might be the cause of this task dependence. It is unclear if it is a conflation between task and the ground truth/displayed statistics concepts discussed earlier, or an actual difference in the way we percieve graphics. A good starting place for this idea would be to identify if there is task dependency in perceptual tasks, since all perceptual task research to date has been done on value retrival. That is, an experiment that compares things like area, position, colour and other perceptual tasks on value exaction, comparisons, and other simple tasks. It is also clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations. A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted. The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error. Of course, this is limited by the fact that there also isnt a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiemnts. @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point. Reguardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.

# Conclusion
In this paper we have highlighted a series of misunderstandings, confusions, and methodical errors in the uncertainty literature.

Many of these issues are not problems in isolation, so long as the authors themselves are aware of it and state these assumptions in their papers. Our problem with this body is work is that the authors themselves seem to be oblivious to this vast array of problems, considering not a single one of the problems discussed here are mentioned explicitly in a single visualisation paper.

>>>>

Unlike uncertainty, the concept of the "information" in a plot is not entirely avoided and works such as *The Grammar of Graphics* attempt to outline exactly what information is contained within a plot. When creating a graphic there are several tasks that must be completed in a specific order, regardless of whether or not we are working with a uncertainty visualisation or not, this pipeline is illustrated in @fig-pipeline, and contains all the steps of the grammar of graphics [@Leland2005]. First the mathematical information in a plot is established in the "varset" setps, and then that mathematical information is visually represented, which is defined in the "graph" steps. Since two plots cannot be compared on their visual elements if they contain different information (because you will never be able to identify if the difference is because of informational asymetry or visual processing) we could understand two plots as being the same if their "varset" steps are identical. However because uncertainty is not defined mathematically, we cannot reliably add it in at the "statistic" stage. Additionally, the grammar of graphics was not designed to extend to "uncertainty visualisations" as @Leland2005 referred to uncertainty as "semantics", so concepts that are common to uncertainty mathematically, such as resampling do not quite fit within the framework. For example, if we have one set of data and two graphics, where one graphic represents the data as a series of points, and the the other depicts a smoothed density using a line. These two plots differ in *both* the statistic and geometry stages of the grammar of graphics, however as n (the number of observations in our sample) gets larger, these two plots will become visually indistinguishable (depending on the collision modifier used). This result aligns nicely with large sample theory, because as n increases the data and mass should become mathematically indistinguishable, and many uncertainty visualisation experiments notice an interaction between sample size and graphic effectiveness so this result is likely also practically true [@Kale2018; @Newburger2022; @Hofmann2012] but despite the increasing similarity in what is actually depicted in the graphic, the sample and mass visualisations would remain different visualisations within the grammar of graphics framework. Additionally, authors rarely attribute this visual version of large sample theory to results where it is clearly applicable. @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition might produce the same result. The issue of blurred elements of the grammar of graphics in the case of distribution visualisations does not end here. Visualisations that depict mass such as a confidence interval, a boxplot, a PDF, a letterbox plot, a violin plot, a histogram, a quantile dot plot, etc, all show some version of the mass of a variable at different resolutions, but whether these graphics differ in something as low down in the grammar pipeline as the variable stage, or later at the geometry stage is unclear.

@@

# Uncertainty is noise, not signal
Uncertainty is noise, but most evaluation experiments measure it as signal. Uncertainty acts as a form of statistical "hedging" for signals found in an analysis. Interviews with experts in statistic back up this primary motivation, as ignoring uncertainty information of often expressed as being similar to fraud or lying [@Hullman2020a; @Manski2020]. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it, if that message is invalid.  When we ask the viewer of a plot to look at data and extract a value, we are asking them to perform inference on that value. There will be noise associated with that answer and that is uncertainty. If we ask direct question about some uncertainty metric, we have turned the uncertainty into signal because that is what the participants are drawing inference on.

@@

### More specific hypothesis
Heuristic checks are useful because they look at unknown pitfalls that might exist in interpretation of current plots. Since the hypothesis for these experiments are usually quite specific, e.g. "do people perceive a an outcome that is within the bar as more likely than one outside it, even if both outcomes are the same distance from the mean?". This means they are less likely to fall into the trap of trying to answer questions that are *far* too broad to be answered with a single experiment (e.g. "is a scatter plot better at showing uncertainty than a box plot"). This work also provides useful insights for experiments by highlights pitfalls participants might fall into when they review the results of evaluation experiments [@Hullman2016]. @Newman2012 found that participants were more likely to view points within the bar as more likely than points outside of the bar in bar charts with error bars. Similar effects have been identified in other types of uncertainty displayed. @Padilla2017 found that points that were on an outcome of an ensemble display were perceived as more likely than points not on an outcome, even when the point that was not on a specific outcome of the ensemble was closer to the mean of the uncertainty distribution. The sine illusion can cause the confidence interval of a smoothed sine curve to seem wider at the peaks than the troughs, causing us to underestimate uncertainty associated with changing values [@Vanderplas2015].

In a similar vein, experiments that verify smaller aspects of plot design might be more useful to the field in the long run because it helps contribute to a larger working theory of "how do we see visualisations". Many visualisation experiments try to compare two plots with several differences, but do not seem to be interested in the mechanisms by which we extract information from visualisations. Small perceptual tasks that seek to answer small but highly relevant questions (for example, if colour hue and colour value can be perceived as a single signal suppressed variable) would be useful to the field.

### Qualitative Studies
Alternatively visualisation research could shift away from the accuracy concept all together ask questions that allow for open ended responses. This method can enlighten authors as to *how* the uncertainty information was used by the participants. @Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", however this qualitative assessment does not seem to have made it into the final paper. @Daradkeh2015 presented participants with ten investment alternatives and asked participants "from among available alternatives, which alternative do you prefer the most", and were asked to think aloud and consider the uncertainty in their decision making. The experimenters goal was to observe and organise the methods people use when making decisions in the face of uncertainty. This study was an excellent example in a useful experimental design. They highlighted the specific aspects of uncertainty that participants typically considered, such as the range of outcomes that are above/below a certain threshold, minimum and maximum values, the risk of a loss, etc, and mapped where in the decision making process participants made these considerations. Data visualisation is commonly utilised as a tool in data exploration, so it is not uncommon for a data analyst to make a plot with only a vague goal and pull out a large number of adjacent observations. This experimental framework could replicate this process.

### Just noticiable signal
It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered. The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework. Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticiable to a human at that level. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation.

![Difference between the null plot vs the uncertainty plot. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ](hypvs.jpeg){#fig-hypvs}

# Great Examples 
Despite the common problems detailed in the previous sections, there is some interesting work in the uncertainty visualisation space. This can come in the form of an uncertainty visualisation that attempts to visualise a typically ignored aspect of uncertainty, or an experiment that avoids the pitfalls detailled in the previous sections.

The literature focus on "quantifiable" uncertainty leaves the variance that occurs at early stages in our analysis that is difficult to quantify ignored and forgotten. Some authors have chosen to focus on these unquantifiable cases and look at expressing more complicated cases of uncertainty.  @Tierney2023 builds upon the tidy data pricipals to allow users to handle missing values. This includes data plots with a missing value "shadow" that allows visualisation authors to identify if the variables used in a plot have any structure in their missing values, which would contribute to uncertainty. Another example of uncertainty that is often ignored is the uncertainty resulting from human choices. Climate scenario uncertainty, shown in @fig-climatescenario, attempts to display the range of climate change outcomes that can result from a range of best and worse case human choices.

![Example of a I will just make my own version of a climate scenario plot, this one is just here from a random site as a place holder](climatescenario.png){fig-climatescenario}

Another visualisation method that has a lot of potential is visualisation of samples. Visualisations that opt to express a signal as a sample rather than an estimate have the potential to supress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments. This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing. These sampling methods can show more of the messiness of the data that sits behind a model. This may not have a detrimental effect on the viewers ability to extract global statistics, as it seems they can be extracted from a visualisation of a sample with ease [Franconeri2021]. Sample visualisations have been used in maps with the pixel map, shown in @fig-pixelmap [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000]. This method has also been adopted by visualisation authors outside of academia as can be seen in the  and the New York Times class mobility figure, shown in @fig-nyt, an animated version of which can be found [here](https://www.nytimes.com/interactive/2018/03/19/upshot/race-class-white-and-black-men.html).

@@

A final method to perform signal supression is simply to visualise the data if it is available and relevant to the uncertainty distribution. An example of this is shown in @fig-census for racial distributions spatially in america, an interactive version of the plot can be found [here](https://edition.cnn.com/interactive/2021/us/census-race-ethnicity-map/). This map shows the typical causes of uncertainty in a spatial model, (e.g. regions where data is sparse, ethnically diverse areas, uneven distribution of points within boundaries, etc) but it avoids the need to create a visualisation with a specific signal in mind. This is the technique typically employed by exploratory data analysis, which means it's lack of a specified signal means there is both *no* uncertainty (since we are technically not performing inference) but in the event we *do* implicitly perform inference, there is some hedging. In this sense, the best uncertainty visualisation you can get without specifying a signal you want to convey is visualising the data itself. The census dot map's addition of interactivity also allows users to zoom in and see the details that caused "uncertainty" in the form of inconsistent colours at lower resolutions when they were zoomed out. This does not mean that visualising raw data instead of implementing sampling techniques is always a valid uncertainty visualisation that will prevent insignificant signal from getting through. @Buja2009 illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the lineup protocol was implemented. However it is simple but effective option that seems to be laregly overlooked by the uncertainty visualisaiton community.

@@

@@

Our first suggestion for future work is a mathematical defintion of uncertainty. There is currently a lot of work that quantifies the uncertainty through bias and variance at various stages of the data analysis pipeline separately. Elements such as imputed data, assumptions, sampling methods, and analyst decisions all have their own independent work, quantification and discussions, but methods to brand between these concepts are few and far between. A conceptual framework that allows us to combine these results would be incredibly beneficial to the field.

The concept of uncertainty should also be formalised within the grammar of graphics. This formalisation would allow uncertainty visualisation authors to have a clear understanding of what is or is not an uncertaitny visualisation. Additionally placing uncertainty visualisation in the framework that is used to understand existing information visualisation research would help authors understand when existing methods can be used to explain their results. Incorperating uncertainty into the grammar of graphics will also give a more precise concept of the information contained within a plot. Other fields of science employ marginal changes when designing experiments to ensure it is well understood *what* aspect of their experiment is contributing to their results, and a better sense of what "marginal" is in the case of uncertainty visualisaion would greatly help the field.

Additionally the experimental practices within uncertainty visualisation need to be less ad hoc and more standardised. If we are going to consider uncertainty as noise, not signal, there needs to be a way to identify this signal supression in an experimental design. As the literature currently exists, there is no way to combine papers to get a meaningful sense of how uncertainty information is understood by a viewer. There is also the possibility that uncertainty visualisation evaluations will need to swap to a qualitative methodology where participants are allowed to freely comment on what they notice in graphics until we establish how the existence of noise can be observed.

If an uncertainty visualisation researcher would prefer to perform experiments rather than formalise methods, there are options there too. It would be interesting to know if any perceptual tasks that can be mapped to two different visual tasks condense into a single dimension when looking for overarching signal in a plot. Alternatively, the task dependency many authors in uncertainty visualisation mention would be a useful direction to consider. It is clear that the the number of potential tasks that can be performed on a visualisation increases with with the number of observations. A single observation is limited to value extraction, two observations can be compared, multiple observations allow for shapes or global statistics to be extracted. The interaction between sample size and task is of particular interest to the uncertainty visualisation community, as uncertainty can be expressed through multiple observations using a sample, or through a single value using an error. Of course, this is limited by the fact that there also isnt a definition for what is a "task" and given the mess created by the lack of formalisation in uncertainty visualisation, it may be wise to formalise that concept before performing these experiemnts. @Amar2005 suggested a taxonomy for information visualisation based on the types of tasks we use visualisations for and suggest 10 "analytical primitives" that we can then map to visualisations, which could be a good starting point. Regardless, these are directions of research would be fruitful to the uncertainty visualisation community even if it appears on the surface to be research that is only beneficial to the "normal" visualisation community.

### 2024-03-22 writingdumpster.qmd
finished rewrite

@@

# Main section outtakes
@Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants. Participants were asked to report in which of the plots was "the blue group furthest to the right" and to provide reasoning from a multiple choice list and grade their certainty. The experiment set up is shown in @fig-right. The participants answers were then compared to a ground truth where the correct plot had a right shifted mean. By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.

>>>>

- These industry visualisation are likely made with a large amount of back and forth between different levels of an organisation, something that is important for good visual communication [@Spiegelhalter2017]. 
- @Boone2018 organised responses from participants by the approach they used and tried to account for participants prior beliefs (the method of eliciting prior beleifs was, however, questionable). 
- @Correll2018 “place your 5 ships on the safest locations on the board.” (minimise danger)
- @Blenkinsop2000 did try to include uncertainty as noise, asking participants to search for a specific outcome (the land classified as grass) in a random . However the participants failing to identify the signal (something that would be expected of an uncertainty plot) was seen as a sign the task was too complicated. While the task in this experiment *was* certainty far too complicated (and poorly communicated) this discussion indicates the authors did not consider the signal suppression to be the goal of the visualisation, and rather an annoying distraction. 

@@

# Main section outtakes
Conclusion
In this paper we have highlighted a series of misunderstandings, confusions, and methodical errors in the uncertainty literature. Many of these issues are not problems in isolation, so long as the authors themselves are aware of it and state these assumptions in their papers. Our problem with this body is work is that the authors themselves seem to be oblivious to this vast array of problems, considering not a single one of the problems discussed here are mentioned explicitly in a single visualisation paper.

It is clear that value retrival and questions with a specific ground truth are not appropriate, as uncertainty visualisation will simply follow the existing visualisation rules. Testing a secondary feature of a plot, that is, how strong a signal is becomes a little more complicated. Line up protocols have been used in adjacent work that look at the strength of the signal in a plot, and this idea of identifying if plots have some "barely noticiable differences" could be utilised.

Specifically, a definition that has a place for all the concepts that currently sit under the "uncertainty" umbrella, explains how these concepts relate to uncertianty, and also allows us to quantify them in a way where they can be combined or visualised. This work would not only benefit uncertainty visualisaiton, but also statistics more broadly.

The problem of trying to assess uncertainty without knowing *why* we are depiting uncertainty has been noticed by other authors. @Spiegelhalter2017 noted we "cannot assess the quality of risk communication unless the objectives are clear".

Does the flip (hypothesis test vs uncertainty visualisation) have mathematical implications in how the plots should be assessed? This implies a visible signal in an uncertainty plot should indicate some divergence from the null.

This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot].

It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*.

Using the most commonly presented methods In the absense of an encapsulating mathematical definition this will not change. That being said, even with a definition that allows us to combine these uncertainties into a single value, it would still be beneficial for authors to develop specialty visualisations for different sources of uncertainty, so analysts can utilise visual tools to understand the uncertainty in their data. still might be desirable to see which stage of the analysis is creating uncertainty in the visualisaiton.

@Hofmann2012 showed two distributions in 20 different visualisations (a lineup protocol) using a jittered sample, a density plot, a histogram, and a box plot and asked participants. Participants were asked to report in which of the plots was "the blue group furthest to the right" and to provide reasoning from a multiple choice list and grade their certainty. The experiment set up is shown in @fig-right. The participants answers were then compared to a ground truth where the correct plot had a right shifted mean. By comparing the results to a ground truth statistic and marking participants as "wrong" or "right", the error from the participants that had an alternative interpretation to the concept of "furthest right" was conflated with the error from a the visualisation choice.
